---
title: "Biofilm Bd Update"
author: "Caitlin Nordheim-Maestas"
format:
  html:
    toc: true
    code-fold: true
    toc-location: left
---

# To-do list/ meeting notes from 2/16/24 meeting:

Part I "15 sites pond water": Done

- We are using TOTAL Bd for part I
- Update barplot: side by side barplot with colors, update labels to pond water with and without microbes, order by site by park to keep consistent across plots
- Visualize a boxplot version too
- Visualize a scatterplot too: day 1 v day 7, color by filter

Part II: "Nine sites field biofilm"

- Biofilm only for analysis, but keep all data in the plot
- Stats: this is a randomized block design, so stats accordingly: zsp~ day + site as a random effect
- Went back and forth on color, I think the final consensus was that biofilm is dark grey and supernatant is lighter gray

Part III: Monolayer-Bd biofilm: Done

- Cut the 15 sites supernatant only experiment
- Use total Bd
- Paired ttest on the difference

Part IV: Cross-experiment comparisons: Is a monolayer biofilm a better inhibitor compared to microbes alone?
- 1: susbet out only nine sites
- 2: calculate the percent Bd lost from day 1 to day 7 per site
- 3: add column for monolayer biofilm or no biofilm
- 4: diff ~ type
- Or just plot them 

Part V: Cross-experiment comparisons: Is a field biofilm a better inhibitor compared to microbes alone?
- 1: susbet out only nine sites and nly 40 micron filtered from part1
- 2: calculate the percent Bd lost from day 1 to day 7 per site
- 3: add column for field biofilm or no biofilm
- 4: diff ~ type
- Or just plot them

```{r}
#| warning: false
#| include: false

# Data wrangling

## read in and clean data
library(tidyverse) # for cleaning and viewing data
library(here) # for importing data
library(ggpubr)
library(rstatix)
library(ggplot2)
library(car)
library(multcomp)
library(nlme)

ns_biofilm_bd <- read.csv(here("data", "nine-sites-biofilm-on-Bd - Sheet1.csv"))
fs_pw_bd <- read.csv(here("data", "fifteen-sites-PW-on-Bd - Sheet1.csv"))
#fs_ml_pw_bd <- read.csv(here("data", "fifteen-sites-PW-on-MLBd - Sheet1.csv")) # cut this experiment
ns_ml_pw_bd <- read.csv(here("data", "nine-sites-PW-on-MLBd - Sheet1.csv"))

# remove controls
fs_pw_bd <- fs_pw_bd %>% filter(site != "sterile MQ")

ns_ml_pw_bd<- ns_ml_pw_bd %>% filter(site != "MQ")

ns_ml_pw_bd<- ns_ml_pw_bd %>% filter(site != "NORTH")

#----  Part I

# data type cleaning
fs_pw_bd$bd_location <- factor(fs_pw_bd$bd_location, levels = c("floating", "adherent"))
fs_pw_bd$filter <- factor(fs_pw_bd$filter,
                          levels = c("40um_filter", "0.22um_filter"))
fs_pw_bd$day <- factor(fs_pw_bd$day, levels = c("1", "7"),
                            labels = c("Day_1", "Day_7"))
fs_pw_bd$site <- factor(fs_pw_bd$site,
                          levels = c("BARN", "CABIN", "NORTH", "GRAMPS", "WEST", "GDPND004", "GDPND005", "GDPND006", "GDPND008", "GDPND009", "PRPND002", "PRPND003", "PRPND004", "PRPND009", "PRPND010", "sterile MQ"))

# Step 1: combine floating and adherent
fs_wider_combined <- fs_pw_bd %>%
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating)

fs_wider_combined <- subset(fs_wider_combined, select = -c(adherent,floating) )

fs_wider_combined <- fs_wider_combined %>% 
  # calculate diff
  pivot_wider(names_from = day, values_from = combined_bd) %>%
  mutate(diff = Day_1 - Day_7)

# Step 2: create subsets for each treatment
fs_wider_combined_40um <- fs_wider_combined %>% 
  filter(filter =="40um_filter")
fs_wider_combined_.22um <- fs_wider_combined %>% 
  filter(filter =="0.22um_filter")

#---- Part II
ns_biofilm_bd$site <- factor(ns_biofilm_bd$site,
                                levels = c("CABIN", "GRAMPS", "WEST", "GDPND005", "GDPND006", "GDPND009", "PRPND004", "PRPND009", "PRPND010"))
ns_biofilm_bd$bd_location <- factor(ns_biofilm_bd$bd_location, levels = c("supernatant", "biofilm"))
ns_biofilm_bd$day <- factor(ns_biofilm_bd$day, levels = c("0", "1", "7"), 
                            labels = c("Day_0", "Day_1", "Day_7")) 

biofilm_only <- ns_biofilm_bd %>% 
  filter(bd_location == "biofilm")

ns_biofilm_wider_combined <-  ns_biofilm_bd %>% 
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = biofilm + supernatant)

ns_biofilm_wider_combined <- subset(ns_biofilm_wider_combined, select = -c(biofilm, supernatant))

ns_biofilm_wider_combined <- ns_biofilm_wider_combined %>% 
  filter(day != "Day_0") %>% 
  # calculate diff
  pivot_wider(names_from = day, values_from = combined_bd) %>%
  mutate(diff = Day_1 - Day_7)

  
#---- Part III
ns_ml_pw_bd$site <- factor(ns_ml_pw_bd$site,
                                levels = c("CABIN", "GRAMPS", "WEST", "GDPND005", "GDPND006", "GDPND009", "PRPND004", "PRPND009", "PRPND010", "control"))
ns_ml_pw_bd$bd_location <- factor(ns_ml_pw_bd$bd_location, levels = c("supernatant", "adherent"))
ns_ml_pw_bd$day <- factor(ns_ml_pw_bd$day, levels = c("1", "7"), 
                            labels = c("Day_1", "Day_7")) 

# Step 1: combine supernatant and adherent
ns_ml_wider_combined <- ns_ml_pw_bd %>%
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + supernatant)

ns_ml_wider_combined <- subset(ns_ml_wider_combined, select = -c(adherent,supernatant) )

ns_ml_wider_combined <- ns_ml_wider_combined %>% 
  # calculate diff
  pivot_wider(names_from = day, values_from = combined_bd) %>%
  mutate(diff = Day_1 - Day_7)

# Step 2: create subsets for each treatment
ns_ml_wider_combined_40um <- ns_ml_wider_combined %>% 
  filter(filter =="40um_filter")
ns_ml_wider_combined_.22um <- ns_ml_wider_combined %>% 
  filter(filter =="0.22um_filter")

```

# Part I: Effect of aquatic environmental microbes on Bd growth - 15 sites

## PI About the data

Wrangled results file name: ["fifteen-sites-PW-on-Bd.csv"](https://docs.google.com/spreadsheets/d/11pCxR2oeprag1EGEAtdYpSmjaPwXD9szq_029p4OB2Q/edit?usp=sharing)

File source: "[Biofilm PCR for R-studio](https://docs.google.com/spreadsheets/d/17QveSH4al14axf-cBsu4JEhgampQEU1E/edit?usp=sharing&ouid=116930126837996341679&rtpof=true&sd=true)" first tab called 15 sites PW on Bd

### Statistical questions:

*NOTE: we need to discuss how we are handling adherent vs floating, I did one test combining and a test on each*

Part I Q1

Is there a difference in the **gain or loss of Bd over 7 days** between the **filter sizes** looking at the **TOTAL BD**

**Answer: There is a significant difference in the change in the total quantity of Bd from Day 1 to Day 7 across the filter types (t = 4.7879, df = 14, p-value = 0.000289)**

This experiment took pond water from 15 East Bay sites in Spring 2022 back to the lab, filtered the pond water to keep or remove the microbes, and added Bd to the pond water. Two replicates per site per treatment, one destructively sampled day 1, and one destructively sampled day 7 for qPCR quantification of Bd in the supernatant and the biofilm itself.

## PI Data visualization

### Renwei barplot remake

```{r}
with_microbes <- "#3c3c3c"
no_microbes <- "#888888"

fs_pw_bd %>%
   pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating) %>% 
ggplot(aes(y= combined_bd, x = site, fill = filter)) + 
    geom_col(position = position_dodge()) +
   #scale_y_log10() +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7")))+
  scale_fill_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
    xlab("Site") +
    ylab("Bd Quantity \n (log transformed zoospore equivalents)") +
  guides(fill=guide_legend(title=""))

# Colors
# can try fun colors suggested: these are from Paul Tol's colorblind friendly palette
with_microbes <- "#999933"
no_microbes <- "#88ccee"

fs_pw_bd %>%
   pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating) %>% 
ggplot(aes(y= combined_bd, x = site, fill = filter)) + 
    geom_col(position = position_dodge()) +
  scale_y_log10() +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7")))+
  scale_fill_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
    xlab("Site") +
    ylab("Bd Quantity \n (zoospore equivalents)") +
  guides(fill=guide_legend(title=""))
```
## Boxplot

```{r}
with_microbes <- "#3c3c3c"
no_microbes <- "#888888"

fs_pw_bd %>%
   pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating) %>% 
ggplot(aes(y= log10(combined_bd), x = filter)) + 
    geom_boxplot(fill = no_microbes) +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7")))+
  scale_fill_manual(values = c(no_microbes)) +
    theme_classic() +
   theme(legend.position = "none",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
  scale_x_discrete (labels= c("40um_filter" = "With Microbes", "0.22um_filter" = "No Microbes")) +
    xlab("Presence of Microbes in Pond Water") +
    ylab("Bd Quantity \n (log transformed zoospore equivalents)")

# can try fun colors suggested: these are from Paul Tol's colorblind friendly palette
with_microbes <- "#999933"
no_microbes <- "#88ccee"

fs_pw_bd %>%
   pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating) %>% 
ggplot(aes(y= log10(combined_bd), x = filter, fill = filter)) + 
    geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
    #scale_y_log10() +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7")))+
  scale_fill_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(legend.position = "none",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
  scale_x_discrete (labels= c("40um_filter" = "With Microbes", "0.22um_filter" = "No Microbes")) +
    xlab("Presence of Microbes in Pond Water") +
    ylab("Bd Quantity \n (log zoospore equivalents)") 
```

# visualize the diff

```{r}
# can try fun colors suggested: these are from Paul Tol's colorblind friendly palette
with_microbes <- "#999933"
no_microbes <- "#88ccee"

fs_wider_combined %>%
ggplot(aes(y= diff, x = filter, fill = filter)) + 
    geom_boxplot() +
    #scale_y_log10() +
  scale_fill_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(legend.position = "none",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
  scale_x_discrete (labels= c("40um_filter" = "With Microbes", "0.22um_filter" = "No Microbes")) +
    xlab("Presence of Microbes in Pond Water") +
    ylab("Bd Reduction \n (Quantity of zoospore equivalents lost over 7 days)") 
```

## Scatterplot

```{r}
with_microbes <- "#999933"
no_microbes <- "#88ccee"

fs_wider_combined %>% 
  ggplot(aes(x = log10(Day_7), y = log10(Day_1), color = filter)) +
  geom_point() +
   scale_color_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(legend.position = "bottom",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
    xlab("Bd Quantity Day 7 \n (log transformed zoospore equivalents)") +
    ylab("Bd Quantity Day 1 \n (log transformed zoospore equivalents)") +
  guides(fill=guide_legend(title=""))
  
```


## PI Stats and assumption testing

Question:

Does the difference in Bd from day 1 to day 7 differ between the two filter types?

The samples are essentially paired by site, so a paired t-test is most appropriate

$H0:μ_{difference in Bd}=0$

Assumptions:

Assumes that the observations from each group represent a random sample from the population. Assumes that the difference of the two observations follow a normal distribution.

### Part I Q1: Paired t-test on the difference in all of the Bd from day 1 to day 7 between the two filters

**There is a significant difference in the change in the total quantity of Bd from Day 1 to Day 7 across the filter types (t = 4.7879, df = 14, p-value = 0.000289)**

```{r}
# Assumptions testing

# check normality of the differences across groups
fs_wider_combined_40um %>% 
 ggqqplot("diff", title = "Untransformed Bd in both floating and floating diff between day 1 and 7 qqPlot") # good enough
shapiro.test(fs_wider_combined_40um$diff) # normal, yay!
hist(fs_wider_combined_40um$diff) # good enough

fs_wider_combined_.22um %>% 
 ggqqplot("diff", title = "Untransformed Bd in both floating and floating diff between day 1 and 7 qqPlot") # gorgeous
shapiro.test(fs_wider_combined_.22um$diff) # normal, yay!
hist(fs_wider_combined_.22um$diff) # good enough
  
```

Stats: 

```{r}
# Step 3: run the paired t-test on the difference
t.test(fs_wider_combined_40um$diff, fs_wider_combined_.22um$diff, paired = TRUE)
```


# Part II: Effect of the aquatic environmental biofilm on Bd growth - "9 sites"

Wrangled results file name: ["nine-sites-biofilm-on-Bd.csv"](https://docs.google.com/spreadsheets/d/1YCU_tCge48MhDw7cMfYAZYokpwAVHtkaY7rfrC5jUDg/edit#gid=0)

File source: "[Biofilm PCR for R-studio](https://docs.google.com/spreadsheets/d/17QveSH4al14axf-cBsu4JEhgampQEU1E/edit?usp=sharing&ouid=116930126837996341679&rtpof=true&sd=true)" first tab called 9 sites Biofilm on Bd

### Statisical questions

Part II Q2: Is there a difference in Bd quantity in the biofilm between day 0, 1, 7?

## PII Data visualization

### Renwei barplot

```{r}
ns_biofilm_bd %>%
ggplot(aes(y= bd_qty, x = site, fill=bd_location)) + 
    geom_col() +
    facet_grid(.~day)+
  theme_classic()+
  scale_fill_manual(values = c("lightgrey", "gray45" )) +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = "bottom") + 
    scale_y_log10() +
    xlab("Site") +
   ylab("Total Bd Quantity Per Well \n (zoospore equivalents)") 
```
## Boxplot

```{r}
biofilm_only %>%
ggplot(aes(y= log10(bd_qty), x = day)) + 
    geom_boxplot(fill = "gray45") +
    theme_classic() +
   theme(legend.position = "none",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
  scale_x_discrete (labels= c("40um_filter" = "With Microbes", "0.22um_filter" = "No Microbes")) +
    xlab("Time (days)") +
    ylab("Bd Quantity \n (log transformed zoospore equivalents)") +
  ggtitle("Amount of Bd incorporated into field-collected biofilm over time")
```

## PII Assumptions testing and Stats

Mix-effects model 

Outcome variable: amount of Bd
Fixed effect: day
Random effect: site

### lme time using nlme
```{r}
biofilm_only <- biofilm_only %>% 
  mutate(log_qty = log(bd_qty))

lme_mod <- lme(log_qty ~ day, random = ~1|factor(site),  data = biofilm_only)
lme_mod # "print" the output
summary(lme_mod) # summary of the output

# ignore day 0?
day1_7 <- biofilm_only %>% 
  filter(day!="Day_0")

lme_mod2 <- lme(log_qty ~ day, random = ~1|factor(site),  data = biofilm_only)
lme_mod2 # "print" the output
summary(lme_mod2) # summary of the output

```

### using Generalized linear mixed model

```{r}
library(lme4)

mod<-glmer(log_qty ~ day + (1|site), data=biofilm_only)
summary(mod) # no p output?
```


# Part III: Effect of microbes in pond water on the Bd-release from mono-strain Bd biofilm - "9 sites" and included biofilm

## PIII About the data

Wrangled results file name: ["nine-sites-PW-on-MLBd.csv"](https://docs.google.com/spreadsheets/d/1I2Y4uf3rP3__Y2p3JLrsTD_9PBUNBYgjY7XG1uOidYA/edit?usp=sharing)

File source: "[Biofilm PCR for R-studio](https://docs.google.com/spreadsheets/d/17QveSH4al14axf-cBsu4JEhgampQEU1E/edit?usp=sharing&ouid=116930126837996341679&rtpof=true&sd=true)" tab called 9 sites PW on MLBd

### Statistical question:

Scientific Q: Is mono-strain Bd biofilm resistant to microbes in pond water

Statistical question:

Is there a difference in the **gain or loss of Bd in supernatant over 7 days** between the **filter sizes**?

**There is a significant difference in the change in the total quantity of Bd from Day 1 to Day 7 across the filter types (t = -6.45, df = 7, p-value = 0.0003)**

## Renwei barplot

```{r}
with_microbes <- "#3c3c3c"
no_microbes <- "#888888"

ns_ml_pw_bd %>% 
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + supernatant) %>% 
ggplot(aes(y= log10(combined_bd), x = site, fill = filter)) + 
    geom_col(position = position_dodge()) +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7")))+
  scale_fill_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
    xlab("Site") +
    ylab("Bd Quantity \n (log transformed zoospore equivalents)") +
  guides(fill=guide_legend(title="")) +
  ggtitle("Mono-layer Bd Biofilm exposed to pond water") 

# Colors
# can try fun colors suggested: these are from Paul Tol's colorblind friendly palette
with_microbes <- "#999933"
no_microbes <- "#88ccee"

ns_ml_pw_bd %>% 
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + supernatant) %>% 
ggplot(aes(y= log10(combined_bd), x = site, fill = filter)) + 
    geom_col(position = position_dodge()) +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7")))+
  scale_fill_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
    xlab("Site") +
    ylab("Bd Quantity \n (log transformed zoospore equivalents)") +
  guides(fill=guide_legend(title="")) +
  ggtitle("Mono-layer Bd Biofilm exposed to pond water") 
```

## Boxplot

```{r}
with_microbes <- "#3c3c3c"
no_microbes <- "#888888"

ns_ml_pw_bd %>%
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + supernatant) %>% 

ggplot(aes(y= log10(combined_bd), x = filter, fill = filter)) + 
    geom_boxplot() +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7")))+
  scale_fill_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(legend.position = "none",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
  scale_x_discrete (labels= c("40um_filter" = "With Microbes", "0.22um_filter" = "No Microbes")) +
    xlab("Presence of Microbes in Pond Water") +
    ylab("Bd Quantity \n (log transformed zoospore equivalents)") +
  ggtitle("Mono-layer Bd Biofilm exposed to pond water") 

# can try fun colors suggested: these are from Paul Tol's colorblind friendly palette
with_microbes <- "#999933"
no_microbes <- "#88ccee"

ns_ml_pw_bd %>%
   pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
 mutate(combined_bd = adherent + supernatant) %>% 
ggplot(aes(y= log10(combined_bd), x = filter, fill = filter)) + 
    geom_boxplot() +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7")))+
  scale_fill_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(legend.position = "none",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
  scale_x_discrete (labels= c("40um_filter" = "With Microbes", "0.22um_filter" = "No Microbes")) +
    xlab("Presence of Microbes in Pond Water") +
    ylab("Bd Quantity \n (log transformed zoospore equivalents)") +
  ggtitle("Mono-layer Bd Biofilm exposed to pond water") 
```

## Scatterplot

```{r}
with_microbes <- "#999933"
no_microbes <- "#88ccee"

ns_ml_wider_combined %>% 
  ggplot(aes(x = log10(Day_7), y = log10(Day_1), color = filter)) +
  geom_point() +
   scale_color_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(legend.position = "bottom",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
    xlab("Bd Quantity Day 7 \n (log transformed zoospore equivalents)") +
    ylab("Bd Quantity Day 1 \n (log transformed zoospore equivalents)") +
  guides(fill=guide_legend(title=""))
  
```

## PIII Stats and assumption testing

Question:

Does the difference in Bd from day 1 to day 7 differ between the two filter types?

The samples are essentially paired by site, so a paired t-test is most appropriate

$H0:μ_{difference in Bd}=0$

Assumptions:

Assumes that the observations from each group represent a random sample from the population. Assumes that the difference of the two observations follow a normal distribution.

```{r}
# Assumptions testing

# check normality of the differences across groups
ns_ml_wider_combined_40um %>% 
 ggqqplot("diff", title = "40 um") # not great
shapiro.test(ns_ml_wider_combined_40um$diff) # NOT normal
hist(ns_ml_wider_combined_40um$diff) # nope

# we gotta transform, but we have negative: west GAINED Bd
ns_ml_wider_combined_40um <- ns_ml_wider_combined_40um %>% 
  mutate(log_diff = log(diff))

ns_ml_wider_combined_40um %>% 
  ggqqplot("log_diff", title = "40 um") # good enough
shapiro.test(ns_ml_wider_combined_40um$log_diff) # woohoo!!
hist(ns_ml_wider_combined_40um$log_diff) # sure

# 0.22 micron

# check normality of the differences across groups
ns_ml_wider_combined_.22um %>% 
 ggqqplot("diff", title = ".22 um") # not great
shapiro.test(ns_ml_wider_combined_.22um$diff) # shapiro says its fine
hist(ns_ml_wider_combined_.22um$diff) # meh

# we gotta transform, but we have negative: west GAINED Bd
ns_ml_wider_combined_.22um <- ns_ml_wider_combined_.22um %>% 
  mutate(log_diff = log(diff))

ns_ml_wider_combined_.22um %>% 
  ggqqplot("log_diff", title = ".22 um") # better?
shapiro.test(ns_ml_wider_combined_.22um$log_diff) # woohoo!!
hist(ns_ml_wider_combined_.22um$log_diff) # sure
  
```

Stats: 

```{r}
# Step 3: run the paired t-test on the difference
t.test(ns_ml_wider_combined_40um$log_diff, ns_ml_wider_combined_.22um$log_diff, paired = TRUE)
```
# Part IV Monolayer vs Water

Part IV: Part 1 vs part 3
- 1: susbet out only nine sites
- 2: calculate the percent Bd lost from day 1 to day 7 per site
- 3: add column for field biofilm or no biofilm
- 4: diff ~ type
- Or just plot them

```{r}
# data wrangling for comparisons across experiments

part1 <- fs_wider_combined %>% 
  filter(site!="NORTH" & site!="BARN" & site!="PRPND002"  & site!="PRPND003" & site!="GDPND004"  & site!="GDPND008") %>% 
  mutate(expt = "pond_water")

pw_no_microbes <- part1 %>% 
  filter(filter == "0.22um_filter") %>% 
  mutate(treatment = "PW 0.22um")

pw_microbes <- part1 %>% 
  filter(filter != "0.22um_filter") %>% 
  mutate(treatment = "PW 40um")

part2 <- ns_biofilm_wider_combined %>% 
  mutate(expt = "field_bd_biofilm") %>% 
  mutate(filter = "0.22um_filter")

fieldbf_no_microbes <- part2 %>% 
  filter(filter == "0.22um_filter") %>% 
  mutate(treatment = "Field bf 0.22")

part3 <- ns_ml_wider_combined %>% 
  mutate(expt = "monolayer_bd_biofilm") 

monolayer_no_microbes <- part3 %>% 
  filter(filter == "0.22um_filter") %>% 
  mutate(treatment = "Monolayer 40um")

monolayer_with_microbes <- part3 %>% 
  filter(filter != "0.22um_filter") %>% 
  mutate(treatment = "Monolayer 0.22um")
  
all_parts<- bind_rows(pw_no_microbes, pw_microbes, fieldbf_no_microbes, monolayer_no_microbes, monolayer_with_microbes ) %>% 
  mutate(percent_loss = (diff/Day_1) * 100)

```

## Boxplot of all parts

```{r}
all_parts %>% 
  ggplot(aes(y= percent_loss, x = treatment)) + 
    geom_boxplot() +
  geom_jitter(alpha = 0.5, width = 0.1) +
  scale_fill_manual(values = c(with_microbes, no_microbes), labels=c('Pond water with microbes', 'Pond water without microbes')) +
    theme_classic() +
   theme(legend.position = "none",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
  scale_x_discrete (labels= c("40um_filter" = "With Microbes", "0.22um_filter" = "No Microbes")) +
    xlab("Treatment") +
    ylab("Percent Reduction of Bd over 7 Days") +
  ggtitle("Which treatments are we comparing?") 
```

## What inhibits Bd the most? Microbes vs monolayer vs field biofilm

## throw it all in a mixed model?

```{r}
lme_mod <- lme(percent_loss ~ treatment, random = ~1|factor(site),  data = all_parts)
lme_mod # "print" the output
summary(lme_mod) # summary of the output
```
## Look at just PW microbes vs monolayer with microbes

```{r}
sub1 <- all_parts %>% 
  filter(treatment == "PW 40um" | treatment == "Monolayer 40um")

lme_mod <- lme(percent_loss ~ treatment, random = ~1|factor(site),  data = sub1)
lme_mod # "print" the output
summary(lme_mod) # summary of the output, there's a diff

#alternatively, pairwise t.test
treat1 <- sub1 %>% 
  filter(treatment == "PW 40um")
treat2 <- sub1 %>% 
  filter(treatment != "PW 40um")

t.test(treat1$percent_loss, treat2$percent_loss, paired = TRUE) # there's a difference
```
## Look at just PW microbes vs field bd without microbes

```{r}
sub1 <- all_parts %>% 
  filter(treatment == "PW 40um" | treatment == "Field bf 0.22")

lme_mod <- lme(percent_loss ~ treatment, random = ~1|factor(site),  data = sub1)
lme_mod # "print" the output
summary(lme_mod) # summary of the output, no diff

#alternatively, pairwise t.test
treat1 <- sub1 %>% 
  filter(treatment == "PW 40um")
treat2 <- sub1 %>% 
  filter(treatment != "PW 40um")

t.test(treat1$percent_loss, treat2$percent_loss, paired = TRUE) # no diff
```

# Part V: 

## EDA's

## PI EDA

Boxplot of Bd quantity across days #1

```{r}
all_parts %>%
  ggplot(aes(x = day, y = log(bd_qty +1), fill = bd_location)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom") +
  ggtitle("15 sites ")

fs_pw_bd %>%
  ggplot(aes(x = day, y = log(bd_qty +1), fill = bd_location)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom") +
  facet_wrap(~filter) +
  ggtitle("15 sites ")

```

Overarching question: Does the amount of Bd in the biofilm or supernatant differ across days?

## P2 

Boxplot of Bd quantity across days #1

```{r}
ns_biofilm_bd %>%
  ggplot(aes(x = day, y = log(bd_qty +1), fill = bd_location)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom") +
  ggtitle("Lab Bd innoculation on East Bay Biofilms - '9 sites' ")

```

Boxplot of Bd quantity across days #2: faceted by Bd location to show trends within Bd location

```{r}
ns_biofilm_bd %>%
  ggplot(aes(x = day, y = log(bd_qty +1), fill = bd_location)) +
  geom_boxplot() +
  theme_classic() +
  facet_wrap(~bd_location) +
  theme(axis.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom") +
  ggtitle(" Lab Bd innoculation on East Bay Biofilms - '9 sites' ")
```

Let's look at the distributions of the Bd in biofilms and supernatant

```{r}
# all data biofilms
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>% #filter out only bd in biofilm
  ggplot(aes(x = log(bd_qty +1))) +
  geom_histogram(position = "identity",  bins=30) +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  ggtitle("Bd in BIOFILMS from 9 site experiment")

# faceted by day
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>% #filter out only bd in biofilm
  ggplot(aes(x = log(bd_qty +1))) +
  geom_histogram(position = "identity",  bins=30) +
  facet_wrap(~day) +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  ggtitle("Bd in BIOFILMS from 9 site experiment")

# all data supernatant
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>% #filter out only bd in biofilm
  ggplot(aes(x = log(bd_qty +1))) +
  geom_histogram(position = "identity",  bins=30) +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  ggtitle("Bd in SUPERNATANT from 9 site experiment")

# faceted by day
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>% #filter out only bd in biofilm
  ggplot(aes(x = log(bd_qty +1))) +
  geom_histogram(position = "identity", bins=30) +
  facet_wrap(~day) +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  ggtitle("Bd in SUPERNATANT from 9 site experiment")

```

Now let's check out the qqPlots of Bd in biofilms and supernatant

```{r}
# whole dataset
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
ggqqplot("log_qty", title = "Bd in biofilm qqPlot")

# per day
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
ggqqplot("log_qty", facet.by = "day", title = "Bd in biofilm qqPlot")

# difference between day 1 and day 7
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7) %>%
  mutate(log_diff = log(diff +1)) %>%
  ggqqplot("log_diff", title = "Bd in biofilm diff between day 1 and 7 qqPlot")

# whole dataset
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
ggqqplot("log_qty", title = "Bd in SUPERNATANT qqPlot")

# per day
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
ggqqplot("log_qty", facet.by = "day", title = "Bd in SUPERNATANT qqPlot")

# difference between day 1 and day 7
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7) %>%
  mutate(log_diff = log(diff +1)) %>%
  ggqqplot("log_diff", title = "Bd in supernatant diff between day 1 and 7 qqPlot")

```

```{r}
#| include: false


# shapiro - wilks test: all Bd biofilm
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  mutate(log_bd_qty = log(bd_qty +1)) %>%
  group_by(day) %>%
  shapiro_test(log_bd_qty) # day 0 not normal

# shapiro - wilks test: difference in day 1 and day 7
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7) %>%
  mutate(log_diff = log(diff +1)) %>%
  shapiro_test(log_diff) # normal

```

```{r}
#| include: false

#### paired t-test between day 1 and 7
d1 <- ns_biofilm_bd %>% 
  filter(bd_location == "biofilm") %>%
  filter(day == "Day_1")

d7 <- ns_biofilm_bd %>% 
  filter(bd_location == "biofilm") %>%
  filter(day == "Day_7")

t.test(d1$bd_qty, d7$bd_qty, paired = TRUE)
```

```{r}
#| include: false


# shapiro - wilks test: all Bd biofilm
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  mutate(log_bd_qty = log(bd_qty +1)) %>%
  group_by(day) %>%
  shapiro_test(log_bd_qty) # all good

# shapiro - wilks test: difference in day 1 and day 7
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7) %>%
  mutate(log_diff = log(diff +1)) %>%
  shapiro_test(log_diff) # normal

```


```{r}
#| include: false

#### paired t-test between day 1 and 7
d1 <- ns_biofilm_bd %>% 
  filter(bd_location == "supernatant") %>%
  filter(day == "Day_1")

d7 <- ns_biofilm_bd %>% 
  filter(bd_location == "supernatant") %>%
  filter(day == "Day_7")

t.test(d1$bd_qty, d7$bd_qty, paired = TRUE)
```

