---
title: "Biofilm Bd Update"
author: "Caitlin Nordheim-Maestas"
format:
  html:
    toc: true
    code-fold: true
    toc-location: left
---

# About the data

## Part I: Effect of aquatic environmental microbes on Bd growth  - 15 sites

Wrangled results file name: ["fifteen-sites-PW-on-Bd.csv"](https://docs.google.com/spreadsheets/d/11pCxR2oeprag1EGEAtdYpSmjaPwXD9szq_029p4OB2Q/edit?usp=sharing)

File source: "[Biofilm PCR for R-studio](https://docs.google.com/spreadsheets/d/17QveSH4al14axf-cBsu4JEhgampQEU1E/edit?usp=sharing&ouid=116930126837996341679&rtpof=true&sd=true)" first tab called 15 sites PW on Bd

### Columns:

-   day (1, 7)
-   site (n=1 per site)
-   filter size
-   bd quantity
-   bd location (floating or floating)

### Potential questions to answer explored here: 

**NOTE Caitlin's judgement call** I combined the floating and floating Bd, and we discussed a paired t-test but 2 variables, so presented in a couple of ways:

Could do a t-test on the DIFFERENCE across days on the filter types!

### Part I Q1: Is there a difference in Bd quantity between the filter types on day 1? - paired t-test

### Part I Q2: Is there a difference in Bd quantity between the filter types on day 7? - paired t-test

### Part I Q3: Is there a difference in Bd quantity between the filter types and across days? - 2 way anova-type test lm

This experiment took pond water from 15 East Bay sites in Spring 2022 back to the lab, filtered the pond water to keep or remove the microbes, and added the pond water to lab-cultured Bd-monoculture-biofilms. Two biofilms per water treatment, one destructively sampled day 1, and one destructively sampled day 7 for qPCR quantification of Bd in the supernatant and the biofilm itself.


## Part II: Effect of the aquatic environmental biofilm on Bd growth - "9 sites"

Wrangled results file name: ["nine-sites-biofilm-on-Bd.csv"](https://docs.google.com/spreadsheets/d/1YCU_tCge48MhDw7cMfYAZYokpwAVHtkaY7rfrC5jUDg/edit#gid=0)

File source: "[Biofilm PCR for R-studio](https://docs.google.com/spreadsheets/d/17QveSH4al14axf-cBsu4JEhgampQEU1E/edit?usp=sharing&ouid=116930126837996341679&rtpof=true&sd=true)" first tab called 9 sites Biofilm on Bd

### Columns:

-   day (0, 1, 7)
-   site (n=1 per site)
-   bd quantity
-   bd location (in biofilm or in supernatant)

### Potential questions to answer explored here:

### Part II Q1: Is there a difference in Bd quantity in the biofilm between day 0, 1, 7? - repeated measures anova or a lme

### Part II Q2: Is there a difference in Bd quantity in the supernatant between day 0, 1, 7? - repeated measures anova or a lme

Discussion points:

How to account for site? Paired tests/ repeated measures? It's not truly repeated measured because not on the same sample, but it is from the same initial water sample? Ancova with random effect for site?

# Data wrangling

```{r}
#| warning: false

## read in and clean data
library(tidyverse) # for cleaning and viewing data
library(here) # for importing data
library(ggpubr)
library(rstatix)
library(ggplot2)

ns_biofilm_bd <- read.csv(here("data", "nine-sites-biofilm-on-Bd - Sheet1.csv"))
fs_pw_bd <- read.csv(here("data", "fifteen-sites-PW-on-Bd - Sheet1.csv"))


# data type cleaning

#  Part I
fs_pw_bd$bd_location <- factor(fs_pw_bd$bd_location, levels = c("adherent", "floating"))
fs_pw_bd$filter <- factor(fs_pw_bd$filter,
                          levels = c("40um_filter", "0.22um_filter"))
fs_pw_bd$day <- factor(fs_pw_bd$day, levels = c("1", "7"),
                            labels = c("Day_1", "Day_7"))
fs_pw_bd$site <- factor(fs_pw_bd$site,
                          levels = c("BARN", "CABIN", "NORTH", "GRAMPS", "WEST", "GDPND004", "GDPND005", "GDPND006", "GDPND008", "GDPND009", "PRPND002", "PRPND003", "PRPND004", "PRPND009", "PRPND010", "sterile MQ"))

#  Part II
ns_biofilm_bd$site <- factor(ns_biofilm_bd$site,
                                levels = c("CABIN", "GRAMPS", "WEST", "GDPND005", "GDPND006", "GDPND009", "PRPND004", "PRPND009", "PRPND010"))
ns_biofilm_bd$bd_location <- factor(ns_biofilm_bd$bd_location, levels = c("supernatant", "biofilm"))
ns_biofilm_bd$day <- factor(ns_biofilm_bd$day, levels = c("0", "1", "7"), 
                            labels = c("Day_0", "Day_1", "Day_7")) 
                            
unique(ns_biofilm_bd$site)
# remove the controls
fs_pw_bd <- fs_pw_bd %>%
  filter(site != "sterile MQ")

# no control for nine site expt

```

# Data visulaization

R recreations of Renwei's excel barplots

## Part I: Effect of aquatic environmental microbes on Bd growth

```{r}
fs_pw_bd %>%
ggplot(aes(y= log10(bd_qty + 1), x = site, fill=bd_location)) + 
    geom_col() +
    facet_grid(day~filter)+
    #theme_classic() +
   theme(axis.text.x = element_text(angle = 90),
          legend.position = "bottom") + 
    xlab("Site") +
  ggtitle("Effect of aquatic environmental microbes on Bd growth")
```

## Part II: Lab Bd innoculation on East Bay Biofilms - "9 sites"

```{r}
ns_biofilm_bd %>%
ggplot(aes(y= log10(bd_qty), x = site, fill=bd_location)) + 
    geom_col() +
    facet_grid(.~day)+
    theme(axis.text.x = element_text(angle = 90),
          legend.position = "bottom") + 
    xlab("Site") +
  ggtitle("Lab Bd innoculation on East Bay Biofilms - 9 sites") 
```



# EDA

## Part I:

Boxplot of Bd quantity across days #1

```{r}
fs_pw_bd %>%
  ggplot(aes(x = day, y = log(bd_qty +1), fill = bd_location)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom") +
  ggtitle("15 sites ")

fs_pw_bd %>%
  ggplot(aes(x = day, y = log(bd_qty +1), fill = bd_location)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom") +
  facet_wrap(~filter) +
  ggtitle("15 sites ")

```


## Part II: Lab Bd innoculation on East Bay Biofilms - "9 sites"

Overarching question: Does the amount of Bd in the biofilm or supernatant differ across days?

Boxplot of Bd quantity across days #1

```{r}
ns_biofilm_bd %>%
  ggplot(aes(x = day, y = log(bd_qty +1), fill = bd_location)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom") +
  ggtitle("Lab Bd innoculation on East Bay Biofilms - '9 sites' ")

```

Boxplot of Bd quantity across days #2: faceted by Bd location to show trends within Bd location

```{r}
ns_biofilm_bd %>%
  ggplot(aes(x = day, y = log(bd_qty +1), fill = bd_location)) +
  geom_boxplot() +
  theme_classic() +
  facet_wrap(~bd_location) +
  theme(axis.title = element_text(size = 12, face = "bold"),
        legend.position = "bottom") +
  ggtitle(" Lab Bd innoculation on East Bay Biofilms - '9 sites' ")
```



Let's look at the distributions of the Bd in biofilms and supernatant

```{r}
# all data biofilms
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>% #filter out only bd in biofilm
  ggplot(aes(x = log(bd_qty +1))) +
  geom_histogram(position = "identity",  bins=30) +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  ggtitle("Bd in BIOFILMS from 9 site experiment")

# faceted by day
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>% #filter out only bd in biofilm
  ggplot(aes(x = log(bd_qty +1))) +
  geom_histogram(position = "identity",  bins=30) +
  facet_wrap(~day) +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  ggtitle("Bd in BIOFILMS from 9 site experiment")

# all data supernatant
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>% #filter out only bd in biofilm
  ggplot(aes(x = log(bd_qty +1))) +
  geom_histogram(position = "identity",  bins=30) +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  ggtitle("Bd in SUPERNATANT from 9 site experiment")

# faceted by day
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>% #filter out only bd in biofilm
  ggplot(aes(x = log(bd_qty +1))) +
  geom_histogram(position = "identity", bins=30) +
  facet_wrap(~day) +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  ggtitle("Bd in SUPERNATANT from 9 site experiment")

```

Now let's check out the qqPlots of Bd in biofilms and supernatant

```{r}
# whole dataset
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
ggqqplot("log_qty", title = "Bd in biofilm qqPlot")

# per day
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
ggqqplot("log_qty", facet.by = "day", title = "Bd in biofilm qqPlot")

# difference between day 1 and day 7
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7) %>%
  mutate(log_diff = log(diff +1)) %>%
  ggqqplot("log_diff", title = "Bd in biofilm diff between day 1 and 7 qqPlot")

# whole dataset
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
ggqqplot("log_qty", title = "Bd in SUPERNATANT qqPlot")

# per day
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
ggqqplot("log_qty", facet.by = "day", title = "Bd in SUPERNATANT qqPlot")

# difference between day 1 and day 7
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7) %>%
  mutate(log_diff = log(diff +1)) %>%
  ggqqplot("log_diff", title = "Bd in supernatant diff between day 1 and 7 qqPlot")

```


# Assumptions testing and Stats

## Part I:  Effect of aquatic environmental microbes on Bd growth

Question:

Does the difference in Bd from day 1 to day 7 differ between the two filter types? 

The samples are essentially paired by site, so a paired t-test is most appropriate

$H0:μ_{difference in Bd}=0$

Assumptions:

Assumes that the observations from each group represent a random sample from the population.
Assumes that the difference of the two observations follow a normal distribution.

Plan: 3 paired t-tests: Simplest: combinine floating and not floating Bd, then more complex: One for floating, one for floating, then let Cherie and Renwei choose which is most appropriate

### Part I Q1: Paired t-test on the difference in all of the Bd from day 1 to day 7 between the two filters

**There is a significant difference in the change in the total quantity of Bd from Day 1 to Day 7 across the filter types (t = 4.7879, df = 14, p-value = 0.000289)**


```{r}
# Step 1: combine floating and floating
fs_wider_combined <- fs_pw_bd %>%
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating) %>%
  #mutate(log_combined_bd_qty = log(combined_bd)) %>%
  select(-adherent, -floating) %>% 
  # calculate diff
  pivot_wider(names_from = day, values_from = combined_bd) %>%
  mutate(diff = Day_1 - Day_7)

# Step 2: create subsets for each treatment
fs_wider_combined_40um <- fs_wider_combined %>% 
  filter(filter =="40um_filter")
fs_wider_combined_.22um <- fs_wider_combined %>% 
  filter(filter =="0.22um_filter")

# Step 3: check normality of the differences across groups
fs_wider_combined_40um %>% 
 ggqqplot("diff", title = "Untransformed Bd in both floating and floating diff between day 1 and 7 qqPlot") # good enough
shapiro.test(fs_wider_combined_40um$diff) # normal, yay!
hist(fs_wider_combined_40um$diff) # good enough

fs_wider_combined_.22um %>% 
 ggqqplot("diff", title = "Untransformed Bd in both floating and floating diff between day 1 and 7 qqPlot") # gorgeous
shapiro.test(fs_wider_combined_.22um$diff) # normal, yay!
hist(fs_wider_combined_.22um$diff) # good enough
  
# Step 3: run the paired t-test on the difference
t.test(fs_wider_combined_40um$diff, fs_wider_combined_.22um$diff, paired = TRUE)

```

### Part I Q2: Paired t-test on the difference in floating Bd from day 1 to day 7 between the two filters

**There is a significant difference in the change in the adherent Bd from Day 1 to Day 7 across the filter types (t = -5.8588, df = 14, p-value = 4.155e-05)**

```{r}
# Step 1: SELECT ONLY adherent BD and calculate the diff between day 1 and 7
fs_adherent_wide <- fs_pw_bd %>%
  filter(bd_location =="adherent") %>% 
  # calculate diff
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7)

# Step 2: create subsets for each treatment
fs_adherent_wide_40 <- fs_adherent_wide %>% 
  filter(filter =="40um_filter")
fs_adherent_wide_.22 <- fs_adherent_wide %>% 
  filter(filter =="0.22um_filter")

# Step 3: check normality of the differences across groups
fs_adherent_wide_40  %>% 
 ggqqplot("diff", title = "Untransformed Bd in both adherent and adherent diff between day 1 and 7 qqPlot") # perf
shapiro.test(fs_adherent_wide_40 $diff) # normal, yay!
hist(fs_adherent_wide_40 $diff) # good enough

fs_adherent_wide_.22 %>% 
 ggqqplot("diff", title = "Untransformed Bd in both adherent and adherent diff between day 1 and 7 qqPlot") # gorgeous
shapiro.test(fs_adherent_wide_.22$diff) # normal, yay!
hist(fs_adherent_wide_.22$diff) # good enough
  
# Step 3: run the paired t-test on the difference
t.test(fs_adherent_wide_.22$diff, fs_adherent_wide_40$diff, paired = TRUE)

```

### Part I Q3: Paired t-test on the difference in FLOATING Bd from day 1 to day 7 between the two filters

**Floating Bd alone is not normally distrubuted, violates the test assumptions**

```{r}
# Step 1: SELECT ONLY floating BD and calculate the diff between day 1 and 7
fs_floating_wide <- fs_pw_bd %>%
  filter(bd_location =="floating") %>% 
  # calculate diff
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7)

# Step 2: create subsets for each treatment
fs_floating_wide_40 <- fs_floating_wide %>% 
  filter(filter =="40um_filter")
fs_floating_wide_.22 <- fs_floating_wide %>% 
  filter(filter =="0.22um_filter")

# Step 3: check normality of the differences across groups
fs_floating_wide_40  %>% 
 ggqqplot("diff", title = "40 um floating Bd") # nope
shapiro.test(fs_floating_wide_40$diff) # NOPE
hist(fs_floating_wide_40$diff) # NOPE 

## could log transform and try again if it is of interest

fs_floating_wide_.22 %>% 
 ggqqplot("diff", title = "0.22 um floating Bd") # gorgeous
shapiro.test(fs_floating_wide_.22$diff) # normal, yay!
hist(fs_floating_wide_.22$diff) # good enough
  
# Step 3: run the paired t-test on the difference
#t.test(fs_floating_wide_.22$diff, fs_floating_wide_40$diff, paired = TRUE)

```



## Part II: Dataset 1

### Q1: does the amount of Bd in the biofilm differ from day 1 to day 7 overall?

*Rationale: Only comparing day 1 and 7 because we only have day 1 and 7 in the next experiment, and I thought it made the most sense to be consistent. Later on I try to ask about all of the days.*

Paired t-test

**Yes, there was significantly more Bd in the biofilm on day 1 than day 7 (paired t-test, p = 7.55e-05). Mention log-transformation somewhere in here.**

Assumptions of paired t-test:

1.  Assumes that the observations from each group represent a *random sample* from the population.\
2.  Assumes that the *difference* of the two observations follow a *normal distribution*.

**Question for Cherie:** In biometry we just check the normality of the x variable in each group, do we need to take the difference here and test that? I did, but not sure if it's necessary

Shapiro-wilks tests

```{r}
# shapiro - wilks test: all Bd biofilm
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  mutate(log_bd_qty = log(bd_qty +1)) %>%
  group_by(day) %>%
  shapiro_test(log_bd_qty) # day 0 not normal

# shapiro - wilks test: difference in day 1 and day 7
ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7) %>%
  mutate(log_diff = log(diff +1)) %>%
  shapiro_test(log_diff) # normal

```

Paired t-test code

```{r}
#### paired t-test between day 1 and 7
d1 <- ns_biofilm_bd %>% 
  filter(bd_location == "biofilm") %>%
  filter(day == "Day_1")

d7 <- ns_biofilm_bd %>% 
  filter(bd_location == "biofilm") %>%
  filter(day == "Day_7")

t.test(d1$bd_qty, d7$bd_qty, paired = TRUE)
```

## Q2: does the amount of Bd in the supernatant differ from day 1 to day 7 overall? Paired t-test

**Yes, there was significantly more Bd in the supernatant on day 1 than day 7 (paired t-test, p = 0.00363). Mention log transformation too**

*Rationale: Only comparing day 1 and 7 because we only have day 1 and 7 in the next experiment, and I thought it made the most sense to be consistent. Later on I try to ask about all of the days.*

Paired t-test

**Yes, there was significantly more Bd in the supernatant on day 1 than day 7 (paired t-test, p = 0.00363). Mention log-transformation somewhere in here.**

Assumptions of paired t-test:

1.  Assumes that the observations from each group represent a *random sample* from the population.\
2.  Assumes that the *difference* of the two observations follow a *normal distribution*.

**Question for Cherie:** In biometry we just check the normality of the x variable in each group, do we need to take the difference here and test that? I did, but not sure if it's necessary

Shapiro-wilks tests

```{r}
# shapiro - wilks test: all Bd biofilm
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  mutate(log_bd_qty = log(bd_qty +1)) %>%
  group_by(day) %>%
  shapiro_test(log_bd_qty) # all good

# shapiro - wilks test: difference in day 1 and day 7
ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  pivot_wider(names_from = day, values_from = bd_qty) %>%
  mutate(diff = Day_1 - Day_7) %>%
  mutate(log_diff = log(diff +1)) %>%
  shapiro_test(log_diff) # normal

```

Paired t-test code

```{r}
#### paired t-test between day 1 and 7
d1 <- ns_biofilm_bd %>% 
  filter(bd_location == "supernatant") %>%
  filter(day == "Day_1")

d7 <- ns_biofilm_bd %>% 
  filter(bd_location == "supernatant") %>%
  filter(day == "Day_7")

t.test(d1$bd_qty, d7$bd_qty, paired = TRUE)
```

## Q3: Is there a difference in Bd in the biofilm between day 0, 1, 7?

Repeated Measures ANOVA ?? Is there a better way to do this?

**Take with a lot of salt, not super confident in this one**

Unfamiliar with this test and package, but at first try, looks like there is a difference in Bd in the biofilm across days. Looks like the pairwise comparisons are that day 0 and 1 are different, day 1 and 7 are different, but 0 and 7 are not for the amount of biofilm (after p adjustment).

Unfamiliar test and function to me, followed this tutorial: https://www.datanovia.com/en/lessons/repeated-measures-anova-in-r/

```{r}

# anova
biofilm_only <- ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  mutate(log_qty = log(bd_qty +1))
res.aov <- anova_test(data = biofilm_only, dv = log_qty, wid = site, within = day)
get_anova_table(res.aov) # ges is the generalized effect size (amount of variability due to the within-subjects factor)

# pairwise comparison: pairwise ttest with bonferroni correction
pwc <- ns_biofilm_bd %>%
  filter(bd_location == "biofilm") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
  pairwise_t_test(
    log_qty ~ day, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc

```

### Q4: Is there a difference in Bd in the supernatant between day 0, 1, 7? Repeated Measures ANOVA ?? Is there a better way to do this?

**Take with a lot of salt, not super confident in this one**

Unfamiliar with this test and package, but at first try, looks like there is a difference in Bd in the biofilm across days. Looks like the pairwise comparisons are that all days are different from each other in the supernatant.

Unfamiliar test and function to me, followed this tutorial: https://www.datanovia.com/en/lessons/repeated-measures-anova-in-r/

```{r}

super_only <- ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  mutate(log_qty = log(bd_qty +1))
res.aov <- anova_test(data = super_only, dv = log_qty, wid = site, within = day)
get_anova_table(res.aov) # ges is the generalized effect size (amount of variability due to the within-subjects factor)

# pairwise comparison: pairwise ttest with bonferroni correction
pwc <- ns_biofilm_bd %>%
  filter(bd_location == "supernatant") %>%
  mutate(log_qty = log(bd_qty +1)) %>%
  pairwise_t_test(
    log_qty ~ day, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc

```


