---
title: "Biofilm Bd Update"
author: "Caitlin Nordheim-Maestas"
format:
  html:
    toc: true
    code-fold: true
    toc-location: left
---
# Experiment:

Bd was incubated with aquatic environmental biofilm collected from one pond at UCSB in a 12-well plate, incubated in one of three media (TB broth, MilliQ control water, and microbe depleted pond water filtered through a  0.22 micron filter) and the biofilm was destructively sampled and Bd was quantified the AE biofilm at day 1, 3, 5, 7.  

# Statistical test

Test: 2-way ANOVA with outcome variable of log-transformed Bd quantity and predictor variables of the day and medium with the AE biofilm and Bd and their interaction (logBd ~ day*medium). The log-transformed Bd quantity data pass the assumptions of an ANOVA.

Predictors:

Day: 1, 3, 5, 7 (as factors)

Medium: medium incubated with Bd and aquatic environmental biofilm, either none (milliQ), adding TB proth, or adding filtered (aka microbe depleted) pond water. 

# Load in data and libraries

```{r}
#| warning: false

## read in and clean data
library(tidyverse) # for cleaning and viewing data
library(here) # for importing data
library(car) # stats tests like Levene's
library(multcomp)
library(gt)
library(modelsummary)
library(emmeans)

fig_3b_raw <- read.csv(here("data", "final_NCOS_2024_reformatted_for_R.xlsx - Fig3B.csv"))

# set up custom theme
myCustomTheme <- function() {
  theme_light() +
    theme(axis.text = element_text(size = 12, family = "Times", color = "black"),
          axis.title.x = element_text(margin = margin(t = 10), size = 14, face = "bold", family = "Times", color = "black"), # Add space between x-axis label and axis
          axis.title.y = element_text(margin = margin(r = 10), size = 14, face = "bold", family = "Times", color = "black"), # Add space between y-axis label and axis
          title = element_text(size = 12, face = "bold", family = "Times"),
          plot.caption = element_text(size = 10, face = "italic", family = "Times"),
          legend.text = element_text(size = 10, family = "Times"), # Increase legend text size
          panel.grid.major.x = element_blank(), # Remove major vertical grid lines
          panel.grid.minor.x = element_blank(), # Remove minor vertical grid lines
          panel.grid.major.y = element_blank(), # Remove major horizontal grid lines
          panel.grid.minor.y = element_blank(), # Remove minor horizontal grid lines
          strip.text = element_text(size = 12, face = "bold", family = "Times", color = "black"), # Set strip text style
          strip.background = element_rect(fill = "white", color = "grey"), # Set strip background to white, # color = "black"
          axis.ticks = element_blank() # Remove x and y axis ticks
    )}
```

# Data clean 

```{r}
# add column for microbes or no
ae <- fig_3b_raw %>% 
  rename(sample_ID = Adherent.sample.ID) %>% 
  
  # add columns for components y/n
  # add column for TB or no
  mutate(TB = case_when(
    str_detect(sample_ID, "TB") ~ "y",
    TRUE ~ "n"
  )) %>% 
  # add column for PW or no
  mutate(PW = case_when(
    str_detect(sample_ID, "PW") ~ "y",
    TRUE ~ "n"
  ))

# control data for ae
ae_control_data <- ae %>% 
  filter(day == 0) %>% 
  dplyr::select(day, adh)

# no day 0 for stats
ae_noday0 <- ae %>% 
  filter(day != 0) %>% 
  mutate(log_adh = log(adh)) # note: no zeroes so not log + 1

# quick check: we want day as a FACTOR
ae_noday0 <- ae_noday0 %>% 
  mutate(day = as.factor(day)) %>% 
# column for medium
mutate(medium = sample_ID)
str(ae_noday0$day)

# set MQ as reference
ae_noday0$sample_ID <- factor(ae_noday0$sample_ID)
ae_noday0$sample_ID <- relevel(ae_noday0$sample_ID, ref = "MQ+AEbiofilm")
```

# EDA

```{r}
ae_summary <- ae %>%
  group_by(day, sample_ID) %>%
  reframe(mean = mean(adh), # calculate the mean
            n = length(adh), # count the number of observations
            df = n - 1, # calculate the degrees of freedom
            sd = sd(adh), # calculate the standard deviation
            se = sd/sqrt(n), # calculate the standard error
          ) %>%
  # add column for TB or no
  mutate(TB = case_when(str_detect(sample_ID, "TB") ~ "y", TRUE ~ "n")) %>%
  # add column for PW or no
  mutate(PW = case_when(str_detect(sample_ID, "PW") ~ "y", TRUE ~ "n"))

ae_summary %>%
  # reorder to match Renwei's plot
  mutate(sample_ID = factor(sample_ID,
                            levels = c("1%TB+AEbiofilm", "MQ+AEbiofilm",
                                       "PW+AEBiofilm",   "Added Bd"  ))) %>%

  ggplot(aes(x = day,
           y = mean,
           color = sample_ID)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean - se, # plot the standard error
                    ymax = mean + se),
                width = 0.1) +
  geom_line() + 
  scale_y_log10(limits = c(1e3, 1e7),
                breaks = c(1e3, 1e4, 1e5, 1e6, 1e7)) +
  # vibes
  labs(x = "Day",
       y = "Bd Quantity in AE Biofilm (ZE/well)",
        color = "Medium with\nAE Biofilm")+ # Title for color legend
  
  scale_color_manual(values = c("1%TB+AEbiofilm"= "#4477AA",
                                "MQ+AEbiofilm" = "#228833",
                                #"Added Bd" = "darkgrey",
                                "PW+AEBiofilm" = "#AA3377"),
                    labels = c("1%TB+AEbiofilm" = "TB Broth",
                               "MQ+AEbiofilm" = "MQ Water",
                               "PW+AEBiofilm" = "Filtered\nPond Water",
                                "Added Bd" = "Initial Bd")) + # Custom labels

 myCustomTheme()+
  scale_x_continuous(breaks = c(0, 1, 3, 5, 7),
                   labels = c("Initial\nBd", "1", "3", "5", "7"))
```

**Current legend:**  Inhibitory effect of aquatic environmental biofilm on Bd growth. Bd was incubated with aquatic environmental biofilm collected from one pond at UCSB in a 12-well plate, incubated in one of three media (TB broth, MilliQ control water, and microbe depleted pond water filtered through a  0.22 micron filter) and the biofilm was destructively sampled and Bd was quantified the AE biofilm at day 1, 3, 5, 7. Scatterplot of Bd quantity in the aquatic environmental biofilm from UCSB. X-axis indicates the duration of Bd in the microorganism-depleted pond water (“Filtered pond water”), or 1%TB or MQ. Y-axis represents the mean Bd quantity of the triplicate data as zoospore equivalents per well, and the bar represents the standard error.


# 2 way anova

```{r}
# build model
mod1 <- aov(log_adh ~ day*medium,
  data = ae_noday0)

# diagnostic plot
par(mfrow = c(2,2))
plot(mod1) # kinda not homoskedastic but Cherie says it is good enough

# look at results
summary(mod1)
AIC(mod1) # 37.78083 better than null

TukeyHSD(mod1) # all days sig diff from each other, all media sig diff from each other

# use emmeans package to get the t value
mod1 <- aov(log_adh ~ day*medium,
  data = ae_noday0)
em <- emmeans(mod1,  ~ day * medium) # pairwise comparisons
em_tukey <- pairs(em, adjust = "tukey") # gives t ratio
em_tukey
```

# Results write up

A two-way ANOVA revealed that there was a statistically significant difference in Bd load across days (F(3, 24) = 110.100, p = p <0.0001), across the media (F(2, 24) = 246.55, p <0.0001), and the interaction between the effects of day and medium were also significant (F(6, 24) = 10.145, p = p <0.0001). Bd was significantly lower with each day (Tukey-Kramer with 95% family-wise confidence level, p <0.005 for all) and TB plus biofilm has most Bd inhibition power, followed by pond water with no microbes, then by milliQ with the least inhibition power (Tukey-Kramer with 95% family-wise confidence level, p <0.05 for all). **what do I even say for the interactions?**

# Big ol' table `emmeans`

```{r}
tukey_df <- as.data.frame(em_tukey)

tukey_df %>%
  dplyr::select(contrast, estimate, SE, df, t.ratio, p.value) %>%
  gt() %>%
  cols_label(
    contrast = "Comparison",
    estimate = "Estimate",
    SE = "Standard Error",
    df = "Degrees of Freedom",
    t.ratio = "t-Ratio",
    p.value = "p-value"
  ) %>%
  tab_header(
    title = "Tukey HSD Post-hoc Test Results"
  ) %>%
  fmt_number(
    columns = vars(estimate, SE, t.ratio),
    decimals = 2
  ) %>%
  fmt_number(
    columns = vars(p.value),
    decimals = 3,
    rows = p.value >= 0.001
  ) %>%
  fmt_scientific(
    columns = vars(p.value),
    decimals = 1,
    rows = p.value < 0.001
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  )


```


# Big ol' table `TukeyHSD`

```{r}
tukey_results <- TukeyHSD(mod1) # all days sig diff from each other, all media sig diff from each other

# Convert Tukey HSD results to data frames
tukey_day <- as.data.frame(tukey_results$day)
tukey_day$comparison <- rownames(tukey_day)

tukey_medium <- as.data.frame(tukey_results$medium)
tukey_medium$comparison <- rownames(tukey_medium)

tukey_interaction <- as.data.frame(tukey_results$`day:medium`)
tukey_interaction$comparison <- rownames(tukey_interaction)

# Combine all results into one data frame
tukey_combined <- rbind(
  data.frame(tukey_day, factor = "Day"),
  data.frame(tukey_medium, factor = "Medium"),
  data.frame(tukey_interaction, factor = "Interaction")
)

tukey_combined <- tukey_combined %>%
  mutate(
    diff = round(diff, 3),
    lwr = round(lwr, 3),
    upr = round(upr, 3)
  )

tukey_combined %>%
  dplyr::select(factor, comparison, diff, lwr, upr, p.adj) %>%
  gt() %>%
  cols_label(
    factor = "Comparison",
    comparison = "Group Comparison",
    diff = "Mean Diff",
    lwr = "Lower CI",
    upr = "Upper CI",
    p.adj = "p-value"
  ) %>%
  tab_header(
    title = "Tukey HSD Post-hoc Test Results for AE Biofilm"
  ) %>% 
  fmt_number(
    columns = vars(p.adj),
    decimals = 3,
    rows = p.adj >= 0.001
  ) %>%
  fmt_scientific(
    columns = vars(p.adj),
    decimals = 1,
    rows = p.adj < 0.001
  )
```

