---
title: "2A: East Bay Pond Water"
author: "Caitlin Nordheim-Maestas"
#format: docx
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    toc-location: left
---

**Statistical question**: Is there a difference in the **rate of gain or loss of Bd over 6 days** between the **filter sizes** looking at the **TOTAL BD**

# Load in data and libraries

```{r}
#| warning: false

## read in and clean data
library(tidyverse) # for cleaning and viewing data
library(gt) # pretty stats tables
library(broom) # cleaning for gt
library(here) # for importing data
library(car) # stats tests like Levene's
library(Hmisc) # autocalculate stat summaries in ggplot
library(multcomp) # stats
library(nlme) # mixed effects models
library(emmeans) # for pairwise comparisons, especially on mixed effects models and glms
library(ggpubr) # for making ggqq plot
library(patchwork) # for combining figures

# load "East Bay" experiments data
ns_biofilm_bd <- read.csv(here("data", "nine-sites-biofilm-on-Bd - Sheet1.csv"))
fs_pw_bd <- read.csv(here("data", "fifteen-sites-PW-on-Bd - Sheet1.csv"))

# Colors: these are from Paul Tol's colorblind friendly palette
with_microbes_40_color <- "#999933"
no_microbes_.22_color <- "#88ccee"

# set up custom theme
myCustomTheme <- function() {
  theme_light() +
    theme(axis.text = element_text(size = 12, family = "Times", color = "black"),
          axis.title.x = element_text(margin = margin(t = 10), size = 14, face = "bold", family = "Times", color = "black"), # Add space between x-axis label and axis
          axis.title.y = element_text(margin = margin(r = 10), size = 14, face = "bold", family = "Times", color = "black"), # Add space between y-axis label and axis
          title = element_text(size = 12, face = "bold", family = "Times"),
          plot.caption = element_text(size = 10, face = "italic", family = "Times"),
          legend.text = element_text(size = 10, family = "Times"), # Increase legend text size
          panel.grid.major.x = element_blank(), # Remove major vertical grid lines
          panel.grid.minor.x = element_blank(), # Remove minor vertical grid lines
          panel.grid.major.y = element_blank(), # Remove major horizontal grid lines
          panel.grid.minor.y = element_blank(), # Remove minor horizontal grid lines
          strip.text = element_text(size = 12, face = "bold", family = "Times", color = "black"), # Set strip text style
          strip.background = element_rect(fill = "white", color = "grey"), # Set strip background to white, # color = "black"
          axis.ticks = element_blank() # Remove x and y axis ticks
    )}
```

# Data Wrangling

```{r}
# remove controls
eb_pw <- fs_pw_bd %>% filter(site != "sterile MQ")

# keep control for labeling plot
eb_pw_controls <- fs_pw_bd %>% 
  filter(site =="sterile MQ") %>% 
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating) %>% 
   mutate(day = case_when(
    day == 1 ~ "Day_1",
    day == 7 ~ "Day_7"))

# data type cleaning
eb_pw$bd_location <- factor(eb_pw$bd_location, levels = c("floating", "adherent"))
eb_pw$filter <- factor(eb_pw$filter,
                          levels = c("40um_filter", "0.22um_filter"))
eb_pw$day <- factor(eb_pw$day, levels = c("1", "7"),
                            labels = c("Day_1", "Day_7"))
eb_pw$site <- factor(eb_pw$site,
                          levels = c("BARN", "CABIN", "NORTH", "GRAMPS", "WEST", "GDPND004", "GDPND005", "GDPND006", "GDPND008", "GDPND009", "PRPND002", "PRPND003", "PRPND004", "PRPND009", "PRPND010", "sterile MQ"))

# get the total difference across days by combining both locations of Bd then subtracting across days
eb_pw_total_diff <- eb_pw %>%
  # combine floating and adherent for total_Bd
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating) %>% 
  subset(select = -c(adherent,floating)) %>% 

# different metrics of difference in Bd
  pivot_wider(names_from = day, values_from = combined_bd) %>%

  # calculate the rate loss by taking the log of each before subtracting
  mutate(rate_loss = log(Day_1) - log(Day_7))

# Split into 2 data frames one for 40 and one for .22
eb_pw_total_diff_40um <- eb_pw_total_diff %>% 
  filter(filter =="40um_filter")
eb_pw_total_diff.22um <- eb_pw_total_diff%>% 
  filter(filter =="0.22um_filter")
```

# EDA

```{r}
# Renwei barplot remake
eb_pw %>%
  # combine floating and adherent for total_Bd
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating) %>% 
ggplot(aes(y= combined_bd, x = site, fill = filter)) + 
    geom_col(position = position_dodge()) +
  scale_y_log10() +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7"))) +
  scale_fill_manual(values = c("40um_filter" = with_microbes_40_color, 
                                "0.22um_filter" = no_microbes_.22_color)) +
    theme_classic() +
   theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
    xlab("Site") +
    ylab("Bd Quantity \n (zoospore equivalents)") +
  guides(fill=guide_legend(title=""))
```

Boxplot

```{r}
# eb_pw %>%
#   # combine floating and adherent for total_Bd
#   pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
#   mutate(combined_bd = adherent + floating) %>% 
#   
#   # create the plot
#   ggplot(aes(y= combined_bd, x = filter, fill = filter)) + 
#     geom_boxplot() +
#   geom_jitter(width = 0.2, alpha = 0.3) +
#     scale_y_log10() +
#     facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
#                                                     "Day_7" = "Day 7")))+
#   scale_fill_manual(values = c("40um_filter" = with_microbes_40_color, 
#                                 "0.22um_filter" = no_microbes_.22_color)) +
#     theme_classic() +
#    theme(legend.position = "none",
#          strip.text = element_text(face="bold"),
#          axis.title = element_text(face = "bold")) + 
#   scale_x_discrete (labels= c("40um_filter" = "With Microbes", "0.22um_filter" = "No Microbes")) +
#     xlab("Presence of Microbes in Pond Water") +
#     ylab("Bd Quantity \n (zoospore equivalents)") +
#   
#   # add controls ad x's
#   geom_point(data = eb_pw_controls, aes(x = filter, y = combined_bd), shape = 4, size = 2)
```

## Stats and assumption testing

Question: Does the difference in Bd from day 1 to day 7 differ between the two filter types?

The samples are essentially paired by site, so a paired t-test is most appropriate

$H0:μ_{difference in Bd}=0$

Assumptions:

Assumes that the observations from each group represent a random sample from the population. Assumes that the difference of the two observations follow a normal distribution.

Assumption testing:

```{r}
# check normality of the differences across groups

# numeric check
eb_pw_40um_shapiro.test <- shapiro.test(eb_pw_total_diff_40um$rate_loss) # normal, yay!
eb_pw_0.22um_shapiro.test <- shapiro.test(eb_pw_total_diff.22um$rate_loss) # normal, yay!

# visual check
eb_pw_40um_qq <- eb_pw_total_diff_40um %>% 
  ggqqplot("rate_loss", title = "40 um")

# Histogram using ggplot2
eb_pw_40um_hist <- eb_pw_total_diff_40um %>% 
  ggplot(aes(x = rate_loss)) +
  geom_histogram(binwidth = 0.2) +
  labs(title = "40 um")

# visual check
eb_pw_0.22um_qq <- eb_pw_total_diff.22um %>% 
  ggqqplot("rate_loss", title = "0.22 um")

# Histogram using ggplot2
eb_pw_0.22um_hist <- eb_pw_total_diff.22um %>% 
  ggplot(aes(x = rate_loss)) +
  geom_histogram(binwidth = 0.2) +
  labs(title = "0.22 um")

eb_pw_40um_qq + eb_pw_40um_hist + eb_pw_0.22um_qq + eb_pw_0.22um_hist

eb_pw_40um_shapiro.test
eb_pw_0.22um_shapiro.test
```

The data for the 40um filter is normally distributed (Shapiro-Wilk test, W = `r round(eb_pw_40um_shapiro.test$statistic, 5)`, p = `r round(eb_pw_40um_shapiro.test$p.value, 4)`), and so is the data for the 0.22 filter (Shapiro-Wilk test, W = `r round(eb_pw_0.22um_shapiro.test$statistic, 5)`, p = `r round(eb_pw_0.22um_shapiro.test$p.value, 4)`).

# Stats

```{r}
# Run the paired t-test on the difference
eb_pw_paired_ttest_result <- t.test(eb_pw_total_diff_40um$rate_loss, eb_pw_total_diff.22um$rate_loss, paired = TRUE)

eb_pw_paired_ttest_result
```

**There is a significant difference in the change in the total quantity of Bd from Day 1 to Day 7 across the filter types (t = `r eb_pw_paired_ttest_result$statistic`, df = `r eb_pw_paired_ttest_result$parameter`, p-value = `r eb_pw_paired_ttest_result$p.value`)**

The t-value is positive, which shows that the first item entered (with microbes) has a larger loss of Bd than the second item entered (no microbes.) Df of 14 is expected, because it's 15 sites.

## Fig 2

```{r}
fig_2a <- eb_pw %>%
  # combine floating and adherent for total_Bd
  pivot_wider(names_from = bd_location, values_from = bd_qty) %>%
  mutate(combined_bd = adherent + floating) %>% 
  
  # create the plot
  ggplot(aes(y= combined_bd, x = filter, fill = filter)) + 
    geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
    scale_y_log10() +
    facet_wrap(~day, labeller = labeller(day = c("Day_1" = "Day 1",
                                                    "Day_7" = "Day 7")))+
  
  scale_fill_manual(values = c("40um_filter" = with_microbes_40_color, 
                                "0.22um_filter" = no_microbes_.22_color)) +
    myCustomTheme() +
   theme(legend.position = "none",
         strip.text = element_text(face="bold"),
         axis.title = element_text(face = "bold")) + 
  scale_x_discrete (labels= c("40um_filter" = "+ AE Microbes", "0.22um_filter" = "- AE Microbes")) +
    xlab("Presence of Microbes in Pond Water") +
    ylab("Bd Quantity (ZE)") +
  
  # add controls ad x's
  geom_point(data = eb_pw_controls, aes(x = filter, y = combined_bd), shape = 4, size = 2)
```

```{r}
# Convert factor day to numeric while preserving original values
pw_noday0$day <- as.numeric(as.character(pw_noday0$day))

fig2B <- pw_summary %>% 
  # reorder to match Renwei's plot
  mutate(sample_ID = factor(sample_ID, 
                            levels = c("1%TB", "MQ", "1%TB+PW+microorganism", "PW+microorganism", "1%TB+PW-microorganism", "PW-microorganism", "Added Bd"))) %>% 
  mutate(day = as.numeric(day)) %>%  
  ggplot(aes(x = day, 
             y = mean, 
             color = sample_ID)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean - se, # plot the standard error
                    ymax = mean + se),
                width = 0.1) +

  # Adding the raw data as a layer with jitter
  geom_point(data = pw_noday0, 
             aes(x = day, 
                 y = adh_plus_sup, 
                 color = sample_ID),  # Raw data points
             position = position_jitter(width = 0.1, seed = 1),
             alpha = 0.3) +
  # add control raw data too
  geom_point(data = pw_control_data, 
             aes(x = day, 
                 y = adh_plus_sup, 
                 color = "#BBBBBB"),  # Raw data points
             position = position_jitter(width = 0.1, seed = 1),
             alpha = 0.3) +
  
  scale_y_log10(limits = c(1e3, 1e8), 
                breaks = c(1e3, 1e4, 1e5, 1e6, 1e7, 1e8)) +
  labs(x = "Day",
       y = "Bd Quantity (ZE)",
       color = "Medium",       # Title for color legend
       linetype = "Microbes\nPresent"  # Title for linetype legend
       ) +
  scale_color_manual(values = c("1%TB" = "#CCBB44", 
                                "MQ" = "#228833", 
                                "1%TB+PW+microorganism" = "#4477AA", 
                                "PW+microorganism" = "#EE6677", 
                                "1%TB+PW-microorganism" = "#66CCEE",
                                #"Added Bd" = "#BBBBBB" # removed bc not really a medium
                                "PW-microorganism" = "#AA3377"),  
                    labels = c("1%TB" = "TB",
                                "MQ" = "MQ",
                                "1%TB+PW+microorganism" = "TB + PW + MO",
                                "PW+microorganism" = "PW + MO",
                                "1%TB+PW-microorganism" = "TB + PW - MO",
                                "PW-microorganism" = "PW - MO",
                                "Added Bd" = "Initial Bd")) + # Custom labels for the color legend
  
    geom_line(aes(linetype = microbes)) + 
  scale_linetype_manual(values = c("n" = "dashed", 
                                   "y" = "solid"),
                        labels = c("n" = "N", "y" = "Y")) +  # Change labels to uppercase N and Y
 myCustomTheme()+
  scale_x_continuous(breaks = c(0, 1, 3, 5, 7),
                   labels = c("Initial\nBd", "1", "3", "5", "7")) +
  theme(legend.position = "right",
        panel.grid.major.y = element_line(color = "grey"), # Add major y grid lines
        panel.border = element_blank())
```

# *Publication figure

```{r}
# Combine fig_2a and fig_2b side by side
fig2 <- fig_2a + fig2B + 
  plot_layout(widths = c(1, 1)) +   # Ensure equal widths for both plots
  plot_annotation(tag_levels = 'A')  # Adds "A" and "B" to the upper corners

fig2

#ggsave("2a_2b.png", plot = fig2, width = 14, height = 5, dpi = 1000)
```


